Shader Pipeline and Effect Encapsulation using CLOS

### 2019-01-30 els2019-shader-pipeline/paper.tex
Fix some stuff. Ready for review

\begin{abstract}
  Modern real-time graphics make use of a lot of tricks in order to produce stunning visuals. Many of these tricks require separate rendering passes, as well as separate rendering logic for each pass. These passes are then combined in a variety of ways in order to produce a final image. The interaction between such a rendering pass and the objects it draws, as well as the interaction between multiple passes within a pipeline can become quite complex. Often times in order to handle this complexity, the passes and objects are generalised, and the render logic is left almost entirely with the object or the pass. We present a new method of representing objects, passes, and pipelines, which allows a modular encapsulation of effects and rendering behaviour, as well as object-oriented composition through inheritance.
\end{abstract}

@@

Each of the customisable stages only accepts a single shader for each pass, making it difficult to separate, encapsulate, and ultimately combine behaviour. Furthermore, the steps required in order to change the shaders and shader inputs can be non-trivial and expensive to execute. \\

@@

The complexity of managing the graphics state and the order of rendering can be massively complicated for modern requirements. Rendering a frame often requires a multitude of passes, each with their own parameters and shared data. The rendering of each object within a pass can also differ, leading to even more state that needs to be correctly managed. \\

@@

Gyrling\cite{fibers} presents an overview of the techniques used to perform parallel rendering in Naughty Dog's commercial engine. Individual stages within a pass, render passes of a frame, and multiple frame renderings are divided up into many small jobs that can run in parallel and are synchronised using counters on a shared structure. \\

The case study of the Unity game engine by Messaoudi et al\cite{unity} shows the availability of a set of fixed rendering pipelines that can be customised in a very limited extent with custom shaders. However, these shaders must fit into Unity's existing lighting and overall rendering model. While Unity does allow building a custom pipeline via their Scriptable Rendering Pipeline\cite{unitycustom}, they do not seem to offer any specific encapsulation or modularity features. \\

@@

\section{Overview}\label{overview}
The system is composed out of three distinct entities: shader objects, shader passes, and pipelines. Both shader objects and shader passes encapsulate rendering logic. The pipeline on the other hand represents the assembly of a complete frame and only influences rendering in the sense that it tracks which passes to run in which order. \\

@@

Finally, in \hyperref[concept]{section 7} we show the results of an implementation of a medium-sized pipeline using these techniques.

\section{Passes}\label{passes}
A shader pass should encapsulate the logic for drawing objects in a certain way. In order to do this, the pass needs to be able to partially or fully control the shaders of objects. For instance, an object should still be able to have control over how its vertices are constructed and what materials and textures are applied to it. The shader pass needs to be able to make use of this information if it needs to, or discard it completely if unneeded. \\

In order to allow this we make use of the shader composition capabilities that we presented in our previous work\cite{glsloop} that allows us to combine pieces of GLSL code, and a new protocol centred around shader passes.

@@

For any object that should be rendered using a given pass, first \code{register-object-for-pass} must be called. This allows the pass to prepare the shader program that will be used during rendering for this object. A shader program is an OpenGL resource that compiles all shaders for a pass together. With \code{shader-program-for-pass} this program can then later be retrieved. The shader program is important for objects to access, as it allows them to set values for ``uniform'' variables that are used in the shader code. \\

@@

Objects are typically rendered using the \code{render} function. Object classes are encouraged to add methods that specialise the behaviour and perform necessary setup, as well as to perform the final draw call. For instance, a very primitive class could look as shown in \autoref{lst:simple-object}.

@@

This function is fine for allowing the object being drawn control over the behaviour. However, the pass cannot exert the same amount of control due to the generic function's argument precedence. We thus introduce another function with inverted argument order that is called first. \\

\code{render-with} thus exists mostly as an entry point for rendering operations to allow shader passes greater control over the rendering. Typically it will simply defer to \code{render}. This inversion is very important for shader passes that take all control away from the objects. Without \code{render-with}, the \code{:before}, \code{:after}, and \code{:around} methods on \code{render} specialised on an object could try to manipulate state before methods specialised on a shader pass could prevent it. \\

@@

For instance, the ``Deferred Lighting'' pass shown in \autoref{fig:render-pipeline} is created by combining the passes illustrated in \autoref{lst:deferred-render}, \autoref{lst:shadow-render}, and \autoref{lst:high-pass} through inheritance. \\

@@

\begin{listing}[h]
\begin{minted}[fontsize=\small]{common-lisp}
(let ((pipeline (make-pipeline))
      (shadow (make-instance 'shadow-map-pass))
      (geometry (make-instance 'geometry-pass))
      (deferred (make-instance 'deferred-lighting-pass))
      (blur (make-instance 'gaussian-blur-pass))
      (composite (make-instance 'composite-pass)))
  (connect (port shadow 'shadow-map) 
           (port deferred 'shadow-map) pipeline)
  (connect (port geometry 'position-map)
           (port deferred 'position-map) pipeline)
  (connect (port geometry 'normal-map)
           (port deferred 'normal-map) pipeline)
  (connect (port geometry 'albedo-map)
           (port deferred 'albedo-map) pipeline)
  (connect (port deferred 'high-pass)
           (port blur 'previous-pass) pipeline)
  (connect (port deferred 'color)
           (port composite 'color) pipeline)
  (connect (port blur 'color)
           (port composite 'bloom) pipeline)
  (prepare pipeline))
\end{minted}
\caption{An assembly of the pipeline shown in \autoref{fig:render-pipeline}.}
\label{lst:pipeline}
\end{listing}

@@

\subsection{Constraint Merging}\label{texture-constraints}
OpenGL textures include a massive amount of information.\cite{texparameter}\cite{teximage2d} When two ports are connected that specify different constraints on the texture properties, a join must be performed. Fortunately, a wide range of the texture property values are fundamentally incompatible, meaning that a lot of the logic can simply error. For simplicity and brevity, we will focus on the join operator for a single texture property here, the internal format. The internal format specifies how many colour channels the texture has, how many bits of precision each channel has, which format each channel has, as well as whether the texture is compressed or has sRGB gamma normalisation applied. \\

The list of specified texture formats is quite large.\cite{texinternalformat} Unfortunately OpenGL does not give us an interface to handle these formats in a way that lets us pick the individual features easily. Instead, each format is represented by a constant number, with no relation to the features that format includes. This means that we need to first destructure each format into a list of features:

@@

If the join is successful, we then re-encode the texture format specification into OpenGL's constant and use this constant in the real texture specification. Note that this join could texture format specifications that are not legal according to the OpenGL specification. However, this can only occur if one of the specs to join is already illegal. We thus deem it unnecessary to handle such cases.

@@

\begin{step}
\item The set of unique texture specifications is computed by joining each port's texspec with every other and comparing for equality.
\item For each unique texture specification T:
  \begin{step}
  \item The nodes are sorted topologically.
  \item The number of colours is set to 0.
  \item For each node N:
    \begin{step}
    \item For each output port P of N:
      \begin{step}
      \item If P's texspec is joinable with T...
      \item The number of colours is increased
      \end{step}
    \end{step}
  \item An array is allocated to fit the number of colours. Each index of the array represents a colour and each value at the index whether the colour is currently available or unavailable.
  \item For each node N in \emph{reverse order}:
    \begin{step}
    \item For each input port P of N:
      \begin{step}
      \item For each neighbour port O of P:
        \begin{step}
        \item If O's texspec is joinable with T...
        \item and O does not yet have a colour...
        \item The first available colour is assigned to O.
        \item This colour is marked as unavailable.
        \end{step}
      \end{step}
    \item For each non-input port P of N:
      \begin{step}
      \item If P's texspec is joinable with T...
      \item and P does not yet have a colour...
      \item The first available colour is assigned to P.
      \item This colour is marked as unavailable.
      \end{step}
    \item For each port P of N:
      \begin{step}
      \item If P's texspec is joinable with T...
      \item and P has a colour...
      \item P's colour is marked as available.
      \end{step}
    \end{step}
  \end{step}
\end{step}

@@

This algorithm is by no means efficient, but since pipeline allocation only has to happen during loading phases, we currently do not consider this to be a big problem. \\

@@

\section{Proof of Concept}\label{concept}
As a proof of concept we have implemented these protocols and mechanisms, as well as the passes shown in \autoref{fig:render-pipeline}. As the code required to do so is quite lengthy and involved, and the graphics techniques used are beyond the scope of this paper, we will omit them. The relevant outline of the passes and the pipeline involved has already been shown in \hyperref[pipelines]{section 5}. \\

@@

\begin{figure}[H]
  \begin{center}
    \begin{minipage}{.2\textwidth}
      \includegraphics[width=1.0\textwidth]{pipeline-position.png}
    \end{minipage}
    \begin{minipage}{.2\textwidth}
      \includegraphics[width=1.0\textwidth]{pipeline-normal.png}
    \end{minipage}
    \vskip0.04cm
    \begin{minipage}{.2\textwidth}
      \includegraphics[width=1.0\textwidth]{pipeline-albedo.png}
    \end{minipage}
    \begin{minipage}{.2\textwidth}
      \includegraphics[width=1.0\textwidth]{pipeline-specular.png}
    \end{minipage}
  \end{center}
  \caption{The output of the geometry pass, producing a ``G-buffer'' composed out of a position-, normal-, albedo-, and specular-map.}
  \label{fig:pipeline-gbuffer}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=0.3\textwidth]{pipeline-shadow.png}
  \end{center}
  \caption{The depth map produced by the shadow map pass. It shows the distance of surfaces from the point of view of the primary light source.}
  \label{fig:pipeline-shadow}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \begin{minipage}{.2\textwidth}
      \includegraphics[width=1.0\textwidth]{pipeline-lighting.png}
    \end{minipage}
    \begin{minipage}{.2\textwidth}
      \includegraphics[width=1.0\textwidth]{pipeline-highpass.png}
    \end{minipage}
  \end{center}
  \caption{The color map (left) and the high-pass map (right) that is output by the deferred render pass. It combines the information from the G-buffer from \autoref{fig:pipeline-gbuffer}, the shadow-map from \autoref{fig:pipeline-shadow}, and information for lights to render an image. Note that due to gamma adjustments the image appears quite dark.}
  \label{fig:pipeline-lighting}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=0.3\textwidth]{pipeline-blur.png}
  \end{center}
  \caption{The blurred high-pass from \autoref{fig:pipeline-lighting} produced by the gaussian blur pass. This is used to produce an effect called bloom in the final composite pass.}
  \label{fig:pipeline-blur}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=0.3\textwidth]{pipeline-frame.png}
  \end{center}
  \caption{The final output frame that is produced by adding together the bloom texture from \autoref{fig:pipeline-blur} and the standard colour texture from \autoref{fig:pipeline-lighting}. Gamma adjustment and High Dynamic Range reduction are used for colour normalisation.}
  \label{fig:pipeline-frame}
\end{figure}

@@

These protocols and algorithms are largely generic enough that they could also be adapted for use in other languages and systems. However, we believe that the Common Lisp Object System gives us a set of very convenient tools to make everything feel more natural and more integrated with the rest of the system than other environments would allow us to. \\

@@

\section{Further Work}\label{further-work}
Currently there exist several severe restrictions in our framework. \\

@@

Another restriction is that there is no way to encapsulate GLSL code other than associating it with a class, which is a rather heavy-weight operation. It would be far more useful to introduce a form of library mechanism that allows the definition of standalone shader functions, which can then be imported by shader classes.

@@

\section{Implementation}\label{implementation}
An implementation of the proposed system can be found at
\\\href{https://github.com/Shirakumo/trial/blob/f34a79f0a6df21d1ed9259e85fbb3c7eed39352b/shader-pass.lisp}{https://github.com/Shirakumo/trial/blob/\\f34a79f0a6df21d1ed9259e85fbb3c7eed39352b/shader-pass.lisp}
\\\href{https://github.com/Shirakumo/trial/blob/f34a79f0a6df21d1ed9259e85fbb3c7eed39352b/pipeline.lisp}{https://github.com/Shirakumo/trial/blob/\\f34a79f0a6df21d1ed9259e85fbb3c7eed39352b/pipeline.lisp}
\\\url{https://github.com/Shinmera/flow} \\

>>>>

\begin{abstract}
  Modern real-time graphics make use of a lot of tricks in order to produce stunning visuals. Many of these tricks require separate rendering passes, as well as separate rendering logic for each pass. These passes are then combined in a variety of ways in order to produce a final image. The interaction between such a rendering pass and the objects it draws, as well as the interaction between multiple passes within a pipeline can become quite complex. Often times, in order to handle this complexity, the passes and objects are generalised, and the render logic is left almost entirely with either the object or the pass. We present a new method of representing objects, passes, and pipelines, which allows a modular encapsulation of effects and rendering behaviour, as well as object-oriented composition through inheritance.
\end{abstract}

@@

Each of the customisable stages only accept a single shader for each pass, making it difficult to separate, encapsulate, and ultimately combine behaviour. Furthermore, the steps required in order to change the shaders and shader inputs can be non-trivial and expensive to execute. \\

@@

The complexity of managing this graphics state and the order of rendering can be massively complicated for modern requirements. Rendering a frame often requires a multitude of passes, each with their own parameters and shared data. The rendering of each object within a pass can also differ, leading to even more state that needs to be correctly managed. \\

@@

Gyrling\cite{fibers} presents an overview of the techniques used to perform parallel rendering in Naughty Dog's private game engine. Individual stages within a pass, render passes of a frame, and multiple frame renderings are divided up into many small jobs that can run in parallel and are synchronised using counters on a shared structure. How rendering logic requirements are communicated is however not explained. \\

The case study of the Unity game engine by Messaoudi et al\cite{unity} shows the availability of a set of fixed rendering pipelines that can be customised in a very limited extent with custom shaders. These shaders must fit into Unity's existing lighting and overall rendering model. While Unity does allow building a custom pipeline via their Scriptable Rendering Pipeline\cite{unitycustom}, they do not seem to offer any specific encapsulation or modularity features. \\

@@

\section{Overview}\label{overview}
The system is composed out of three distinct entities: shader objects, shader passes, and pipelines. Both shader objects and shader passes encapsulate rendering logic. The pipeline on the other hand represents the assembly of a complete frame and only influences rendering in the sense that it tracks which passes to run in what order. \\

@@

Finally, in \hyperref[concept]{section 7} we show the results of implementing a medium-sized pipeline using these techniques.

\section{Passes}\label{passes}
A shader pass should encapsulate the logic for drawing objects in a certain way. In order to do this, the pass needs to be able to partially or fully control the shaders of objects. For instance, an object should still be able to have control over how its vertices are constructed and what materials and textures are applied to it. The shader pass should be able to make use of this information if it needs to, or discard it completely if unneeded. \\

In order to allow this we make use of the shader composition capabilities that we presented in our previous work\cite{glsloop}. This technique allows us to combine pieces of GLSL code. On top of this shader combination sits a new protocol to control the interaction between shader passes and objects. \\

@@

For any object that should be rendered using a given pass, first \code{register-object-for-pass} must be called. This allows the pass to prepare the shader program that will be used during rendering for this object. A shader program is an OpenGL resource that compiles the shaders for the stages of a pass together. Using \\ \code{shader-program-for-pass} this program can then later be retrieved. The shader program is important for objects to access, as it allows them to set values for ``uniform'' variables that are used in the shader code. \\

@@

Objects are typically rendered using the \code{render} function. Users are encouraged to add methods that specialise the behaviour and perform necessary setup, as well as to perform the final draw call for their custom object classes. For instance, a very primitive class could look as shown in \autoref{lst:simple-object}. \\

@@

This function is fine for allowing the objec control over the behaviour. However, the pass cannot exert the same amount of control due to the generic function's argument precedence. We thus introduce another function with inverted argument order that is called first. \\

\code{render-with} thus exists mostly as an entry point to allow shader passes greater control over the rendering. Typically it will simply defer to \code{render}. This inversion is very important for shader passes that take all control away from the objects. Without \code{render-with}, the \code{:before}, \code{:after}, and \code{:around} methods on \code{render} specialised on an object could try to manipulate state before methods specialised on a shader pass could prevent it. \\

@@

For instance, the ``Deferred Render'' pass shown in \autoref{fig:render-pipeline} is created by combining the passes illustrated in \autoref{lst:deferred-render}, \autoref{lst:shadow-render}, and \autoref{lst:high-pass} through inheritance. \\

@@

\begin{listing}[h]
\begin{minted}[fontsize=\small]{common-lisp}
(let ((pipeline (make-pipeline))
      (shadow (make-instance 'shadow-map-pass))
      (geometry (make-instance 'geometry-pass))
      (deferred (make-instance 'deferred+shadow-pass))
      (blur (make-instance 'gaussian-blur-pass))
      (composite (make-instance 'composite-pass)))
  (connect (port shadow 'shadow-map) 
           (port deferred 'shadow-map) pipeline)
  (connect (port geometry 'position-map)
           (port deferred 'position-map) pipeline)
  (connect (port geometry 'normal-map)
           (port deferred 'normal-map) pipeline)
  (connect (port geometry 'albedo-map)
           (port deferred 'albedo-map) pipeline)
  (connect (port deferred 'high-pass)
           (port blur 'previous-pass) pipeline)
  (connect (port deferred 'color)
           (port composite 'color) pipeline)
  (connect (port blur 'color)
           (port composite 'bloom) pipeline)
  (prepare pipeline))
\end{minted}
\caption{An assembly of the pipeline shown in \autoref{fig:render-pipeline}.}
\label{lst:pipeline}
\end{listing}

@@

\subsection{Constraint Merging}\label{texture-constraints}
OpenGL textures include a massive amount of information.\cite{texparameter}\cite{teximage2d} When two ports are connected that specify different constraints on the texture properties, a join must be performed. Fortunately, a wide range of the texture property values are fundamentally incompatible, meaning that a lot of the logic can simply error. For simplicity and brevity, we will focus on the join operator for a single texture property here, the internal format. The internal format is arguably the most important property. This property specifies how many colour channels the texture has, how many bits of precision each channel has, which format each channel has, as well as whether the texture is compressed or has sRGB gamma normalisation applied. \\

The list of specified texture formats is quite large.\cite{texinternalformat} Unfortunately OpenGL does not give us an interface to handle these formats in a way that lets us pick the individual features easily. Instead, each format is represented by a constant, whose value has no relation to the features that format includes. This means that we need to first destructure each format name into a list of features: \\

@@

If the join is successful, we then re-encode the texture format specification into OpenGL's constant and use this constant in the real texture specification. Note that this join could produce texture format specifications that are not legal according to the OpenGL specification. However, this can only occur if one of the specifications to join is already illegal. We thus deem it unnecessary to handle such cases.

@@

\begin{step}
\item The set of unique texture specifications is computed by joining each port's texture specification with every other and comparing for equality.
\item For each unique texture specification T:
  \begin{step}
  \item The nodes are sorted topologically.
  \item The number of colours is set to 0.
  \item For each node N:
    \begin{step}
    \item For each output port P of N:
      \begin{step}
      \item If P's texture specification is joinable with T...
      \item The number of colours is increased
      \end{step}
    \end{step}
  \item An array is allocated to fit the number of colours. Each index of the array represents a colour and each value at the index whether the colour is currently available or unavailable.
  \item For each node N in \emph{reverse order}:
    \begin{step}
    \item For each input port P of N:
      \begin{step}
      \item For each neighbour port O of P:
        \begin{step}
        \item If O's texture specification is joinable with T...
        \item and O does not yet have a colour...
        \item The first available colour is assigned to O.
        \item This colour is marked as unavailable.
        \end{step}
      \end{step}
    \item For each non-input port P of N:
      \begin{step}
      \item If P's texture specification is joinable with T...
      \item and P does not yet have a colour...
      \item The first available colour is assigned to P.
      \item This colour is marked as unavailable.
      \end{step}
    \item For each port P of N:
      \begin{step}
      \item If P's texture specification is joinable with T...
      \item and P has a colour...
      \item P's colour is marked as available.
      \end{step}
    \end{step}
  \end{step}
\end{step}

@@

This algorithm is by no means efficient, but since pipeline allocation only has to happen during loading phases, we currently do not consider this to be a big problem. We also have not performed any analysis as to whether the algorithm produces optimal allocation results in every case.\\

@@

\section{Proof of Concept}\label{concept}
As a proof of concept we have implemented these protocols and mechanisms, as well as the passes shown in \autoref{fig:render-pipeline}. As the code required to do so is quite lengthy and involved, and the graphics techniques used are beyond the scope of this paper, we will omit the code. The relevant outline of the passes and the pipeline involved has already been shown in \hyperref[pipelines]{section 5}. \\

@@

\begin{figure}[h]
  \begin{center}
    \begin{minipage}{.2\textwidth}
      \includegraphics[width=1.0\textwidth]{pipeline-position.png}
    \end{minipage}
    \begin{minipage}{.2\textwidth}
      \includegraphics[width=1.0\textwidth]{pipeline-normal.png}
    \end{minipage}
    \vskip0.04cm
    \begin{minipage}{.2\textwidth}
      \includegraphics[width=1.0\textwidth]{pipeline-albedo.png}
    \end{minipage}
    \begin{minipage}{.2\textwidth}
      \includegraphics[width=1.0\textwidth]{pipeline-specular.png}
    \end{minipage}
  \end{center}
  \caption{The output of the geometry pass, producing a ``G-buffer'' composed out of a position- (top left), normal- (top right), albedo- (bottom left), and specular-map (bottom right).}
  \label{fig:pipeline-gbuffer}
\end{figure}

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.3\textwidth]{pipeline-shadow.png}
  \end{center}
  \caption{The depth map produced by the shadow map pass. It shows the distance of surfaces from the point of view of the primary light source.}
  \label{fig:pipeline-shadow}
\end{figure}

\begin{figure}[h]
  \begin{center}
    \begin{minipage}{.2\textwidth}
      \includegraphics[width=1.0\textwidth]{pipeline-lighting.png}
    \end{minipage}
    \begin{minipage}{.2\textwidth}
      \includegraphics[width=1.0\textwidth]{pipeline-highpass.png}
    \end{minipage}
  \end{center}
  \caption{The color map (left) and the high-pass map (right) that is output by the deferred render pass. It combines the information from the G-buffer from \autoref{fig:pipeline-gbuffer}, the shadow-map from \autoref{fig:pipeline-shadow}, and information for lights to render an image. Note that due to gamma adjustments the image appears quite dark.}
  \label{fig:pipeline-lighting}
\end{figure}

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.3\textwidth]{pipeline-blur.png}
  \end{center}
  \caption{The blurred high-pass from \autoref{fig:pipeline-lighting} produced by the gaussian blur pass. This is used to produce an effect called bloom in the final composite pass.}
  \label{fig:pipeline-blur}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{pipeline-frame.png}
  \end{center}
  \caption{The final output frame that is produced by adding together the bloom texture from \autoref{fig:pipeline-blur} and the standard colour texture from \autoref{fig:pipeline-lighting}. Gamma adjustment and High Dynamic Range reduction are used for colour normalisation.}
  \label{fig:pipeline-frame}
\end{figure}

@@

These protocols and algorithms are largely generic enough that they could also be adapted for use in other languages and systems. However, we believe that the Common Lisp Object System gives us a set of very convenient tools to make everything feel more natural and more integrated with the rest of the system than other environments would allow us to do. \\

@@

\section{Further Work}\label{further-work}
Currently there exist several restrictions in our framework. \\

@@

Another restriction is that there is currently no way to encapsulate GLSL code other than associating it with a class, which is a rather heavy-weight operation. It would be far more useful to introduce a library mechanism that allows the definition of standalone shader functions, which can then be required by shader classes.

@@

\section{Implementation}\label{implementation}
An implementation of the proposed system can be found at
\\\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/shader-pass.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/shader-pass.lisp}
\\\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/pipeline.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/pipeline.lisp}
\\\url{https://github.com/Shinmera/flow} \\
